---
title: "Plots and tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plots and tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(champs)
library(tidyverse)
d <- load_data("../_ignore/CHAMPS_de_identified_data/")
# d <- load_data("_ignore/CHAMPS_de_identified_data/")
```

The champs package comes with a few plot and html table functions to help present the output of some of the `calc_` functions.

```r
library(champs)
library(tidyverse)
d <- load_data("CHAMPS_de_identified_data")
```

## Plot functions

Plot methods only exist for a subset of `calc_` functions. Currently, the following are implemented:

- `plot_syndrome_combinations()` which works with the output from `calc_syndrome_combinations()`
- `plot_top_etiol_by_age()` which works with the output from `calc_top_etiol_by_age()`
- `plot_margins_site_age()` which works with the output from `calc_cc_allcases_by_site_age()`, ` calc_detected_allcases_by_site_age()`, and `calc_cc_detected_by_site_age()`.
- `heatmap_site_age()` which works with the same `calc_` functions as `plot_margins_site_age()`.

### Plot syndrome combinations for a specified condition

```{r}
d_syndrome <- calc_syndrome_combinations(d,
  condition = "Streptococcus pneumoniae",
  syndrome_names = c(
    "Lower respiratory infections",
    "Meningitis/Encephalitis",
    "Neonatal sepsis",
    "Congenital infection"),
  syndrome_values = c(
    "Pneumonia",
    "Meningitis",
    "Sepsis",
    "Sepsis"),
  specimen_types <- c(
    "Cerebrospinal fluid sample",
    "Tissue specimen from lung",
    "Whole blood")
)

plot_syndrome_combinations(d_syndrome, legend_location = "none", 
  plot_type = "bar", include_text = "full_percent")

plot_syndrome_combinations(d_syndrome, legend_location = "bottom", 
  include_text = "percent")

plot_syndrome_combinations(d_syndrome, "age_breakdown", legend_location = "bottom", 
include_text = "number")

plot_syndrome_combinations(d_syndrome, "tac_age_breakdown", 
  legend_location = "none", include_text = "full_number")

plot_syndrome_combinations(d_syndrome, "age_breakdown", legend_location = "bottom", 
  include_text = "percent", full_separator = ",")

plot_syndrome_combinations(d_syndrome, "age_breakdown", plot_type = "pie", 
  legend_location = "left", include_text = "number", full_separator = ",")

plot_syndrome_combinations(d_syndrome, "tac_age_breakdown", plot_type = "bar", 
  legend_location = "none", include_text = "full_number", full_separator = "\n")  
```

### Plot top etiologies by specified age groups

```{r}
d_top_etiol_by_age <- calc_top_etiol_by_age(d, age_groups = c(
  "Death in the first 24 hours",
  "Early Neonate (24-72 hours)",
  "Early Neonate (72+hrs to 6 days)",
  "Late Neonate (7 to 27 days)"))

plot_top_etiol_by_age(d_top_etiol_by_age)
```

### Plot margins by site and age

The `plot_margins()` function works with all the `calc_` functions that create numerator and denominator tables with age group and site tabulations.

```{r,fig.width=10}
specimen_types <- c(
  "Cerebrospinal fluid sample",
  "Tissue specimen from lung",
  "Whole blood",
  "Rectal swab",
  "Plasma or spun blood specimen")

d_dcd <- calc_cc_allcases_by_site_age(d, condition =  "Streptococcus agalactiae")
d_tac <- calc_detected_allcases_by_site_age(d, condition =  "Streptococcus agalactiae", pathogen = "Group B Streptococcus")
d_both <- calc_cc_detected_by_site_age(d,
  condition = "Streptococcus agalactiae",
  pathogen = "Group B Streptococcus",
  specimen_types = specimen_types)

plot_margins_site_age(d_tac, include_x_label = FALSE, include_text = "full_fraction",
  plot_title = "TAC Marginal Distributions for Streptococcus agalactiae")

plot_margins_site_age(d_dcd, include_x_label = TRUE, include_text = "fraction",
  plot_title = "DeCoDe Marginal Distribution for Streptococcus agalactiae")             

plot_margins_site_age(d_both, include_x_label = FALSE, include_text = "full_fraction",
  plot_title = "DeCoDe Streptococcus agalactiae where TAC contains Streptococcus agalactiae")

plot_margins_site_age(d_tac, include_x_label = FALSE, include_text = "full_percent", 
  plot_title = "TAC marginal distributions for Streptococcus agalactiae")

plot_margins_site_age(d_tac, include_x_label = FALSE, include_text = "name",
  plot_title = "TAC Marginal Distributions for Streptococcus agalactiae")

plot_margins_site_age(d_dcd, include_x_label = FALSE, include_text = "full_fraction",
  plot_title = "DeCoDe Marginal Distribution for Streptococcus agalactiae")
             
plot_margins_site_age(d_dcd, include_x_label = TRUE, include_text = "fraction",
  plot_title = "DeCoDe Marginal Distribution for Streptococcus agalactiae")             
```

### Plot table percentages heatmap

The `heatmap_site_age()` function works with all the `calc_` functions that create numerator and denominator tables using the age group and site tabulations. A visualized table where each cell is colored by the percentage based on the numerator and denominator tables from the `calc_` functions is output.

```{r}
heatmap_site_age(d_tac, include_text = "percent",
  plot_title = "TAC Results", plot_subtitle = "Streptococcus agalactiae")

heatmap_site_age(d_dcd, include_text = "fraction",
  plot_title = "DeCoDe results", plot_subtitle = "Streptococcus agalactiae")             

heatmap_site_age(d_both, include_text = "fraction",
  plot_title = "DeCoDe Streptococcus agalactiae where TAC contains Streptococcus agalactiae")
```

### Plot function details

The plot functions are built using [ggplot2](https://ggplot2.tidyverse.org/) from the [tidyverse](https://www.tidyverse.org/). They leverage three ggplot2 extension packages to improve the labeling of the pie and bar charts.

- [ggforce](https://ggforce.data-imaginist.com/) provides access to `geom_arc_bar()` for improved pie charts.
- [ggfittext](https://wilkox.org/ggfittext/) provides improved automatic labeling in bar charts.
- [cowplot](https://wilkelab.org/cowplot/articles/plot_grid.html) provides dual plot view for the marginal plots.

The plotting outputs are not intended to be highly customizable, but you can modify much of the output with additional `ggplot2` commands. 

## Table functions

The `html_table_site_age()` function works on `calc_cc_allcases_by_site_age()`, `calc_detected_allcases_by_site_age()`, and `calc_cc_detected_by_site_age()` tabulations. 

```{r}
html_table_site_age(d_dcd)

html_table_site_age(d_dcd, margin_denom = TRUE)

html_table_site_age(d_dcd, margin_denom = TRUE, inside_pct = TRUE)

html_table_site_age(d_dcd, margin_denom = TRUE, inside_denom = TRUE)
```

```r
head(mock$dmg)
#>   champs_deid site_iso_code site  age_group age_group_subcat location_of_dea…
#>   <chr>       <chr>         <chr> <fct>     <fct>            <chr>           
#> 1 HH23B       V             Vbrgm Early Ne… Early Neonate (… facility        
#> 2 SA73Z       G             Gtmpo Early Ne… Early Neonate (… facility        
#> 3 HS54M       W             Wdurc Death in… Death in the fi… facility        
#> 4 KL97A       N             Nugtw Stillbir… Stillbirth       facility        
#> 5 QA17R       W             Wdurc Early Ne… Early Neonate (… facility        
#> 6 ME83B       W             Wdurc Stillbir… Stillbirth       facility        
#> # … with 12 more variables: hosp_los_24h <dbl>, hosp_los_48h <dbl>,
```
